A missing guardrail: The failure of protest risk to deter election manipulation in non-democracies

### 2023-07-10 Output/Submission files/manuscript_anon_v2.Rmd
Edits to election fraud choice and repression

In summary, fraudulent elections are not likely to pose a significant protest risk to incumbents. While opposition actors may seize upon fraudulent elections to initiate a protest, the relationship between this decision and the level of fraud is likely to be weak. Opposition groups are also incentivized to stage protests regardless of the level of fraud, given the highly salient moment afforded by the election itself [@cunhaSignalReceivedAuthoritarian2022]. Fraud itself provides little information about regime strength, and even if a protest movement begins after the election, any useful information provided by the level of fraud is quickly replaced by signals from the protests themselves [@lohmannDynamicsInformationalCascades1994]. Protesters, bystanders, and elites can update their estimation of the relative strength of the protest coalition and the incumbent government by observing the size of crowds, the number of protests, and the government's reaction.

The number and size of protests will instead be related to other indicators of opposition and incumbent capacity. When opposition groups and civil society capacity are more organized and better resourced, it is likely that more protesters will take to the streets in more numerous events (see Chenoweth et al [-@chenowethStateRepressionNonviolent2017] for a review of factors making protest movements more resilient to repression). By contrast, when the government appears well-resourced and resolved, more 'passionate economists' should stay home. As identified in prior research, informative factors should be the previous repressive history of the government [@Davenport2019], the institutionalization of opposition parties [@trejo_ballot_2014], the margin of victory in the election [@simpser2013], and whether or not the chief executive faces term limits [@hale_patronal_2014]. Generally speaking, these structural variables indicate aspects of the political opportunity structure that make protest more likely to occur [@meyer_protest_2004].

@@

However, the theory articulated above does *not* expect that the severity of fraud will be associated with the course of protest. Instead, it predicts that underlying indicators of opposition and ruling-party strength will influence the direction of protest movements. In particular, it expects the size and number of protest events will be sensitive to the repressive capacity of the state and to indicators of opposition strength. that high and low levels of repression will be associated with smaller protest movements (in number and size).

>H5: The relationship between repressive capacity and the number of protest events and the maximum event size is curvilinear, with number and size likely to to first increase and then fall as repressive capacity grows.

>H6: The number of protest events and the maximum event size are likely to increase with indicators of opposition and civil-society strength.

@@

To capture the severity of *election fraud*, I use the V-Dem measure of intentional election irregularities, a category which includes "double IDs, intentional lack of voting materials, ballot-stuffing, misreporting of votes, and false collation of votes." I operationalize electoral manipulation using this variable, rather than aggregate measures of election integrity, because fraud of this kind is thought to be most likely to trigger protest [@Linebarger2020]---it is often incumbents' tool of last resort [@Sjoberg2016], it can fail to signal strength [@Harvey2020], and it can be more damaging to legitimacy than pre-election manipulation [@birch2011;@Szakonyi2021]. This measure of election integrity should thus serve as a strong test of the protest-oriented model. The variable is used as a dependent variable in the first set of models that test H1, and as a potential explanatory variable in the subsequent models of protest number, size, and duration.

@@

The first set of models test Hypothesis 1, on the relationship between protest risk and fraud. Protest risk is operationalized in three ways. First, the *number of pre-election protests* in ECAV is taken as *prima facie* evidence of a more turbulent and risky election environment\textemdash it indicates greater mobilization capacity in civil society and/or the opposition, and may also indicate an elevated sense of grievance. Second, the repressive capacity of the regime is modeled using the lagged *physical integrity index* for the country-year. In one approach, higher values represent more permissive environments, and should be associated with greater protest risk (and thus reduced fraud). Alternatively, in keeping with the 'murder in the middle' framework [@feinLifeIntegrityViolationsDemocracy1995], the highest protest risks (and most reduced fraud) might be found at middle values of repression; regimes with the most demonstrated repressive capacity deter potential challengers from taking to the streets in the first place, and the least repressive states defuse protest through accommodation [@pierskallaProtestDeterrenceEscalation2010;@mooreRepressionDissentSubstitution2000]. I measure repressive capacity using the *physical integrity* index from V-Dem, lagged one year. Higher values for this variable indicate greater freedom from torture and political killing. I also include the variable's square term to account for the 'murder in the middle' framework. The physical integrity variables are expected to be explanatory in models of protest, but are important control variables in the models of election fraud. Since repressive capacity helps influence protest in expectation [@Ritter2016], incumbents that are more confident of deterring protest in advance should be more willing to engage in fraud under the protest-deterrent model of manipulation.

@@

@@

Finally, several binary control variables are included. First, a variable from NELDA is used to indicate *presidential* elections, since these may include both higher levels of manipulation [@simpser2013] and a greater risk of protest. Next, a variable from NELDA indicates if the chief executive is *term limited* during the election at hand. This variable is included since incumbents running up against constitutional term limits can pose major challenges for the unity of the regime, possibly leading to fractured manipulation efforts and post-election protest [@hale_patronal_2014]. Finally, I include dummy variables that provide a categorical evaluation of state capacity. These are taken from the ordinal version of the V-Dem variable "state source of fiscal revenue" (v2stfisccap). If the V-Dem data codes a state as not capable of raising revenue or primarily relying on loans and foreign aid, I code it as a *weak state*. If the state is described as primarily relying on resource rents and expropriation, I code it is a *rentier state*. States that rely primarily on taxes on property or economic transactions are coded as *tax states*. This distinction is relevant since access to natural resource rents may both make protest less likely and repression more effective [@Girod2018]. Lastly, since strong non-democratic governments may be able to adjust the electoral calendar strategically (to affecting both election manipulation and protest risk), I include a variable from NELDA that indicates *off-schedule elections*.   Table \@ref(tab:summary-table) shows summary statistics for all of these variables. To address concerns about multicollinearity, I present models in the appendix (page 25) that exclude control variables associated with pre-election protest.

>>>>

@@

In summary, fraudulent elections are not likely to pose a significant protest risk to incumbents. While opposition actors may seize upon fraudulent elections to initiate a protest, the relationship between this decision and the level of fraud is likely to be weak. Opposition groups are also incentivized to stage protests regardless of the level of fraud, given the highly salient moment afforded by the election itself [@cunhaSignalReceivedAuthoritarian2022]. Fraud itself provides little information about regime strength, and even if a protest movement begins after the election, any useful information provided by the level of fraud is quickly replaced by signals from the protests themselves [@lohmannDynamicsInformationalCascades1994]. Protesters, bystanders, and elites can update their estimation of the relative strength of the protest coalition and the incumbent government by observing the size of crowds, the number of protests, and the government's reaction. Thus, I expect that Hypothesis 5\textemdash which might be assumed under other manipulation-oriented models of protest\textemdash to *not* be supported.

>H5: The number and maximum size of protest events are likely to increase as the severity of election fraud increases.

Once protests begin, the number and size of protest events is likely to be influenced by the organizational capacity of civil society and the opposition on the one hand, and on the ability of regime elites to remain united on the other. When opposition groups and civil society capacity are more organized and better resourced, it is likely that more protesters will take to the streets in more numerous events (see Chenoweth et al [-@chenowethStateRepressionNonviolent2017] for a review of factors making protest movements more resilient). By contrast, when the government appears well-resourced and resolved, more 'passionate economists' should stay home. Much of this balance is mediated by political and social elites' decision(s) to remain supportive of the incumbent; protests can create division among the incumbent elite (e.g. among hardliners and softliners [@schmitter1986transitions]) and incentives for disaffected elites to move against the incumbent leadership [@kimPathways2019]. In turn, the opportunity created by a divided elite\textemdash and their resources [@robertson_protest_2010]\textemdash can keep protest movements alive.

Since information on elite and mass preferences can be difficult to observe outside liberal democracies, in part due to preference falsification [@Kuran1991], it is difficult to measure the risk of elite splits *a priori*. However, Hale [-@hale_patronal_2014] makes the case that elites become incentivized to turn against incumbent governments when they expect the chief executive's time in office has become limited. While the incumbent's age and popularity may be signals of such, a constitutional term limit can also serve this role. As such, Hypothesis 6 predicts that protest number and size will increase for elections where the incumbent faces a term limit. This hypothesis serves as useful counter-point to the traditional view on election manipulation. Unlike election fraud (as presented here), a constitutional term limit creates a clear, unambiguous signal that enables coordination among protesters and elites that can influence the course of protest.

>H6: The number of protest events and the maximum event size are likely to increase following elections where the chief executive is.

@@

To capture the severity of *election fraud*, I use the V-Dem measure of intentional election irregularities, a category which includes "double IDs, intentional lack of voting materials, ballot-stuffing, misreporting of votes, and false collation of votes." I operationalize electoral manipulation using this variable, rather than aggregate measures of election integrity, because fraud of this kind is thought to be most likely to trigger protest [@Linebarger2020]---it is often incumbents' tool of last resort [@Sjoberg2016], it can fail to signal strength [@Harvey2020], and it can be more damaging to legitimacy than pre-election manipulation [@birch2011;@Szakonyi2021]. This measure of election integrity should thus serve as a strong test of the protest-oriented model. It also matches the conception of election manipulation used in much of the relevant formal literature.^[For example, election manipulation is treated as post-hoc adjustment of the results in Little et al [-@Little2015], Little [-@little_elections_2012], Magaloni [-@magaloni2010], and Fearon [-@fearon_self-enforcing_2011], among others. Egorov and Sonin [-@Egorov2021] is an exception, which focuses on pre-election restrictions.] The variable is used as a dependent variable in the first set of models that test H1, and as a potential explanatory variable in the subsequent models of protest number, size, and duration. As an additional check in the appendix, a fuller measure of overall election integrity is used instead.

@@

The first set of models test Hypothesis 1, on the relationship between protest risk and fraud. Protest risk is operationalized in three ways. First, the *number of pre-election protests* in ECAV is taken as *prima facie* evidence of a more turbulent and risky election environment\textemdash it indicates greater mobilization capacity in civil society and/or the opposition, and may also indicate an elevated sense of grievance.

@@

The number and size of protests will instead be related to other indicators of opposition and incumbent capacity.  As identified in prior research, informative factors should be the previous repressive history of the government [@Davenport2019], the institutionalization of opposition parties [@trejo_ballot_2014], the margin of victory in the election [@simpser2013], and whether or not the chief executive faces term limits [@hale_patronal_2014]. Generally speaking, these structural variables indicate aspects of the political opportunity structure that make protest more likely to occur [@meyer_protest_2004].

@@

Several binary control variables are included. First, a variable from NELDA is used to indicate *presidential* elections, since these may include both higher levels of manipulation [@simpser2013] and a greater risk of protest. Next, a variable from NELDA indicates if the chief executive is *term limited* during the election at hand. This variable is included since incumbents running up against constitutional term limits can pose major challenges for the unity of the regime, possibly leading to fractured manipulation efforts and post-election protest [@hale_patronal_2014]. Finally, I include dummy variables that provide a categorical evaluation of state capacity. These are taken from the ordinal version of the V-Dem variable "state source of fiscal revenue" (v2stfisccap). If the V-Dem data codes a state as not capable of raising revenue or primarily relying on loans and foreign aid, I code it as a *weak state*. If the state is described as primarily relying on resource rents and expropriation, I code it is a *rentier state*. States that rely primarily on taxes on property or economic transactions are coded as *tax states*. This distinction is relevant since access to natural resource rents may both make protest less likely and repression more effective [@Girod2018]. Lastly, since strong non-democratic governments may be able to adjust the electoral calendar strategically (to affecting both election manipulation and protest risk), I include a variable from NELDA that indicates *off-schedule elections*.

Lastly, the repressive capacity of the state deserves extended mention, since a state with greater capacity for repression may be better able to deter protest and thus 'get away with' more manipulation [@Egorov2021,@Ritter2016]. Omitting a control for repressive capacity would likely produce a biased finding, in which high fraud was associated with low levels of protest as a result. I measure repressive capacity using the *physical integrity* index from V-Dem, lagged one year. Higher values for this variable indicate greater freedom from torture and political killing. In one approach, higher values represent more permissive environments, and should be associated with greater protest risk (and thus reduced fraud). Alternatively, in keeping with the 'murder in the middle' framework [@feinLifeIntegrityViolationsDemocracy1995], the highest protest risks (and most reduced fraud) might be found at middle values of repression; regimes with the most demonstrated repressive capacity deter potential challengers from taking to the streets in the first place, and the least repressive states defuse protest through accommodation [@pierskallaProtestDeterrenceEscalation2010;@mooreRepressionDissentSubstitution2000].  I thus include the variable's square term to account for the 'murder in the middle' framework.

Table \@ref(tab:summary-table) shows summary statistics for all of these variables. To address concerns about multicollinearity, I present models in the appendix (page 25) that exclude control variables associated with pre-election protest.
